Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2014-12-10 parts/def-ph.tex
add TODOs after review
more explanations in section PH

@@

@@

In the scope of this paper, we focus on a specific class of PH models
in which actions of $\PHh^{(1)}$ are called \emph{prioritised actions} and allow to model non-biological actions.
We consider in this section a PH model with $k$ classes of priorities: $\PH = (\PHs; \PHl; \PHa^{\langle k \rangle})$, with $k \in \sNN$
and we define here the restrictions that lead to this class of models.

\pref{cr:bounded} states that the dynamics of the studied model $\PH$ contains no infinite sequence of prioritised actions.
As these actions can be considered as non-biological and therefore instantaneous, we thus prevent the existence of any Zeno-like behaviour
which would allow the play of an infinite sequence of prioritised actions in “zero time”.
As a consequence, the set $\restriction{\Sce}{1}$ is finite.

@@

Given a sort $a\in\Sigma$ and a state $s\in L$, 
we denote $\pfp_s(a)$ (\pref{def:pfp}) the processes of sort $a$ that can be present after
playing all actions of priority $1$,
which is always defined because of \pref{cr:bounded}.

@@

Our second condition (\pref{cr:compcs}) imposes that any sort in the Process
Hitting is either a well-formed component (\pref{def:component}) or a well-formed cooperative sort
(\pref{def:cs}).
The former is a sort that is hit only by actions with priority greater than $2$.
The latter is a sort that is hit only by actions with priority $1$ and which
always converges to the same process with respect to the state of the components
that have an influence on it (\pref{def:conn}).
We note $\components\subset\Sigma$ the set of well-formed components,
and $\cs\subset\Sigma$ the set of well-formed cooperative sorts.

@@

@@

We note that a sort $a$ that is hit by no action (\ie $\forall h \in \PHh, \PHsort(\target{h}) \neq a$)
does comply with both definitions of well-formed components and well-formed cooperative sorts.
In this case, such a sort can be arbitrarily affected to either set.
Such a case may occur for “input” sorts (whose state is initially fixed and does not evolve)
or some special cases of cooperative sorts produced by the translations of \pref{sec:encodings}.

@@

In the following we simply write “component” (\resp “cooperative sort”) instead of “well-formed component” (\resp “well-formed cooperative sort”).

\begin{example}
  The PH in \pref{fig:ph-livelock} contains three components ($a$, $b$ and $c$) and a cooperative sort ($ab$, which models cooperation between sorts $a$ and $b$).
\end{example}

@@

We first denote by $\update(s)$ the state equivalent to $s$ but in which all cooperative sorts are updated (\pref{def:update}).
This state is unique due to the properties of $\pfp$ given in the previous subsection.
Then, \pref{lem:update} states that from any state, there exists a scenario updating the cooperative sorts of this state.
%
\begin{definition}[$\update : \PHl \rightarrow \PHl$]
\label{def:update}
  For all $s \in \PHl$, we define:
  \begin{align*}
    \update(s) = s \Cap \{ \pfp_{s}(a) \mid a \in \cs \} \enspace.
  \end{align*}
\end{definition}
%
\begin{lemma}
\label{lem:update}
  $\forall s \in \PHl, \exists \delta \in \restriction{\Sce}{1}(s), s \PHplay \delta = \update(s)$
\end{lemma}
%
\begin{proof}
  Let $a$ be a cooperative sort so that $\PHget{s}{a} \neq \pfp_s(a)$.
  Given the definition of $\pfp_s(a)$, there exists a scenario $\delta$ updating $a$ in $s$ so that
  $\forall \delta' \in \restriction{\Sce}{1}(s \PHplay \delta)$, $\PHget{(s \PHplay \delta \PHplay \delta')}{a} = \pfp_s(a)$.
\end{proof}

\pref{lem:hcompcomp} states that for a given state $s$, and for any action $h = \PHhit{a_i}{b_j}{b_k}$ where $a$ and $b$ are components,
if $\PHget{s}{a} = a_i$ and $\PHget{s}{b} = b_j$, then
$h$ can always be played after a series of prioritised hits (and these hits do not prevent it to be fired).
\pref{lem:hcscomp} states the same if $a$ is a cooperative sort, under the condition that $a$ is updated in $s$.
\begin{lemma}
\label{lem:hcompcomp}
  $\forall s \in \PHl, \forall a,b \in \components, \forall h = \PHhit{a_i}{b_j}{b_k} \in \PHh,$\\
  $(\PHget{s}{a} = a_i \wedge \PHget{s}{b} = b_j) \Rightarrow (\exists \delta \in \restriction{\Sce}{1}(s),
  (s \PHplay \delta) \PHPtrans (s \PHplay \delta \PHplay h))$
\end{lemma}
%
\begin{proof}
  From \pref{lem:update}, there exists a scenario $\delta$ with: $(s \PHplay \delta) = \update(s)$.
  As $a,b \in \components$, $a_i \in (s \PHplay \delta)$ and $b_j \in (s \PHplay \delta)$.
  Finally, by definition of $\update(s)$, no prioritised action can be played in $(s \PHplay \delta)$, thus $h$ can be played in $(s \PHplay \delta)$.
\end{proof}
%
\begin{lemma}
\label{lem:hcscomp}
  $\forall s \in \PHl, \forall h = \PHhit{a_i}{b_j}{b_k} \in \PHh, a \in \cs, b \in \components,$\\
  $(\PHget{s}{a} = a_i \wedge \PHget{s}{b} = b_j \wedge \pfp_s(a) = a_i) \Rightarrow (\exists \delta \in \restriction{\Sce}{1}(s),
  (s \PHplay \delta) \PHPtrans (s \PHplay \delta \PHplay h))$
\end{lemma}
\begin{proof}
  Similar to the proof of \pref{lem:hcompcomp};
  as $a_i \in \pfp_s(a)$, $a_i \in (s \PHplay \delta)$.
\end{proof}

>>>>

\todo{Gluer : explication des définitions, etc.}

@@

\todo{Rajouter un exemple}

@@

In the scope of this paper, we focus on a specific class of PH models
in which actions of $\PHh^{(1)}$, called \emph{prioritised actions},
are only used to model cooperations in the form of strict logic gates.
\modMF{%
For this, we formally define in this section the notion of well-formed cooperative sorts,
and we require that prioritized actions can only be used to update them.
This particular form in the actions allows to apply the static analysis
developed in \pref{sec:sa}, which is not directly possible in the general case.
However, we note here that this class of models is not less expressive than
the general PH framework as given in \pref{def:ph};
indeed, the translation given in \pref{ssec:flattening} proves that
this class of models can reproduce any desired dynamics of the general case.
}%

We consider in this section a PH model with $k$ classes of priorities: $\PH = (\PHs; \PHl; \PHa^{\langle k \rangle})$, with $k \in \sNN$
and we define here the restrictions that lead to this class of models.

\pref{cr:bounded} states that the dynamics of the studied model $\PH$ contain no infinite sequence of prioritised actions.
As these actions can be considered as non-biological and therefore instantaneous, we thus prevent the existence of any Zeno-like behaviour
which would allow the play of an infinite sequence of prioritised actions in “zero time”.
As a consequence, the set $\restriction{\Sce}{1}$ is finite.

@@

\modMF{%
\begin{example}
  The model of \pref{fig:ph-livelock} respects \pref{cr:bounded}:
  indeed, whatever are the active processes of $a$ and $b$,
  it is never possible to indefinitely play actions of $\PHh^{(1)}$.
\end{example}
}%

Given a sort $a\in\Sigma$ and a state $s\in L$, 
we denote $\pfp_s(a)$ (\pref{def:pfp}) the processes of sort $a$ that can be present after
playing all actions of priority $1$,
which is always defined because of \pref{cr:bounded}.
\modMF{%
This definition will be useful for cooperative sorts,
in order to determine their local stable state in every configuration,
and thus which configuration each of their processes represent.
}%

@@

\modMF{%
\begin{example}
  In the model of \pref{fig:ph-livelock}, we have for instance:
  $\pfp_{s}(ab) = \{ab_{10}\}$ for the state $s = \PHstate{a_1, b_0, ab_{00}, z_0}$.
  Indeed, starting from $\PHstate{a_1, b_0, ab_{00}, z_0}$,
  it is only possible to play the action $\PHfrappe{a_1}{ab_{00}}{ab_{10}} \in \PH^{(1)}$,
  and no other action in $\PH^{(1)}$ is playable in the resulting state
  $\PHstate{a_1, b_0, ab_{10}, z_0}$.
  Thus, process $ab_{10}$ can be used to represent the configuration $\PHstate{a_1, b_0}$.
\end{example}
}%

Our second condition (\pref{cr:compcs}) imposes that any sort in the Process
Hitting is either a well-formed component (\pref{def:component}) or a well-formed cooperative sort
(\pref{def:cs}).
The former is a sort that is hit only by actions with priority greater than $2$,
\modMF{%
and thus stands for a regular biological component.
}%
The latter is a sort that is hit only by actions with priority $1$ and which
always converges to the same process with respect to the state of the components
that have an influence on it (\pref{def:conn}).
\modMF{%
Indeed, we require that each process of a cooperative sort corresponds
to exactly one configuration of the sorts it represents.
}%
We note $\components\subset\Sigma$ the set of well-formed components,
and $\cs\subset\Sigma$ the set of well-formed cooperative sorts.

@@

\modMF{%
\begin{example}
  For the PH model of \pref{fig:ph-livelock}, we have:
  \begin{align*}
    \conn(a) &= \{a\} &
    \conn(ab) &= \{a, b, ab\} \\
    \conn(b) &= \{b\} &
    \conn(z) &= \{z\}
  \end{align*}
  Therefore, $\compin(ab) = \{a, b\}$.
\end{example}
}%

@@

\modMF{%
\begin{example}
  The sorts $a$, $b$ and $z$ of \pref{fig:ph-livelock} are well-formed components
  regarding \pref{def:component}.
  Furthermore, $ab$ of is a well-formed cooperative sort,
  with respect to \pref{def:cs}, which models cooperation between $a$ and $b$
  because $\compin(ab) = \{a, b\}$ and:
  \begin{enumerate}[(i)]
    \item it is only hit by actions in $\PH^{(1)}$;
    \item in every state, it has exactly one focal process;
    \item each of its processes corresponds to at least one state of the model;
    \item its focal processes only depend on $a$ and $b$.
  \end{enumerate}
  Therefore, we obviously have:
  $\components = \{a, b, z\}$ and $\cs = \{ab\}$,
  which respects \pref{cr:compcs}.
\end{example}
}

We note that any sort $a$ that is hit by no action (\ie $\forall h \in \PHh, \PHsort(\target{h}) \neq a$)
does comply with both definitions of well-formed components and well-formed cooperative sorts.
In this case, such a sort can be arbitrarily affected to either set.
Such a case may occur for “input” sorts (whose state is initially fixed and does not evolve)
or some special cases of cooperative sorts produced by the translations of \pref{sec:encodings}.
In the following we simply write “component” (\resp “cooperative sort”) instead of “well-formed component” (\resp “well-formed cooperative sort”).

@@

\modMF{%
\begin{example}
  We have, for the cooperative sort $ab$ of \pref{fig:ph-livelock}
  $\csState(ab_{10}) = \{\{a_1, b_0\}\}$,
  which is a direct consequence of the fact that $\pfp_{s}(ab) = \{ab_{10}\}$
  with $s = \PHstate{a_1, b_0, ab_{00}, z_0}$.
\end{example}
}%

@@

We first denote by $\update(s)$ (\pref{def:update}) the state equivalent to $s$
\modMF{%
but in which the active process of all cooperative sorts is changed to the process
that represent the current state of the components.
}%
This state is unique due to the properties of $\pfp$ given in the previous subsection.
Then, \pref{lem:update} states that from any state, there exists a scenario updating the cooperative sorts of this state,
\modMF{%
thus allowing to end in $\update(s)$.
}%

\begin{definition}[$\update : \PHl \rightarrow \PHl$]
\label{def:update}
  For all $s \in \PHl$, we define:
  \begin{align*}
    \update(s) = s \Cap \{ \pfp_{s}(a) \mid a \in \cs \} \enspace.
  \end{align*}
\end{definition}
%
\begin{lemma}
\label{lem:update}
  $\forall s \in \PHl, \exists \delta \in \restriction{\Sce}{1}(s), s \PHplay \delta = \update(s)$
\end{lemma}
%
\begin{proof}
  Let $a$ be a cooperative sort so that $\PHget{s}{a} \neq \pfp_s(a)$.
  Given the definition of $\pfp_s(a)$, there exists a scenario $\delta$ updating $a$ in $s$ so that
  $\forall \delta' \in \restriction{\Sce}{1}(s \PHplay \delta)$, $\PHget{(s \PHplay \delta \PHplay \delta')}{a} = \pfp_s(a)$.
\end{proof}

\modMF{%
\begin{example}
  Regarding the model of \pref{fig:ph-livelock}, we have:
  $\update(s) = \PHstate{a_1, b_0, ab_{10}, z_0}$,
  with $s = \PHstate{a_1, b_0, ab_{00}, z_0}$.
  We note that in $\update(s)$, only the active process of the cooperative sort
  $ab$ is changed
  to the process representing the presence of both $a_1$ and $b_0$.
  Furthermore, \pref{lem:update} states that there exists a scenario $\delta$
  of actions in $\PH^{(1)}$
  so that $s \play \delta = \update(s)$.
  Indeed, $\delta = (\PHfrappe{a_1}{ab_{00}}{ab_{10}})$ is such a scenario,
  here containing only one action.
\end{example}
}%

\pref{lem:hcompcomp} states that for a given state $s$, and for any action $h = \PHhit{a_i}{b_j}{b_k}$ where $a$ and $b$ are components,
if $\PHget{s}{a} = a_i$ and $\PHget{s}{b} = b_j$, then
$h$ can always be played after a series of prioritised hits (and these hits do not prevent it to be fired).
\pref{lem:hcscomp} states the same if $a$ is a cooperative sort, under the condition that $a$ is updated in $s$.
In other words, prioritized actions cannot prevent a non-prioritized action
to be eventually played, they only ``delay'' it,
provided only that the hitter of this action
matches the active process in the updated state.

\begin{lemma}
\label{lem:hcompcomp}
  $\forall s \in \PHl, \forall a,b \in \components, \forall h = \PHhit{a_i}{b_j}{b_k} \in \PHh,$\\
  $(\PHget{s}{a} = a_i \wedge \PHget{s}{b} = b_j) \Rightarrow (\exists \delta \in \restriction{\Sce}{1}(s),
  (s \PHplay \delta) \PHPtrans (s \PHplay \delta \PHplay h))$
\end{lemma}
%
\begin{proof}
  From \pref{lem:update}, there exists a scenario $\delta$ with: $(s \PHplay \delta) = \update(s)$.
  As $a,b \in \components$, $a_i \in (s \PHplay \delta)$ and $b_j \in (s \PHplay \delta)$.
  Finally, by definition of $\update(s)$, no prioritised action can be played in $(s \PHplay \delta)$, thus $h$ can be played in $(s \PHplay \delta)$.
\end{proof}
%
\begin{lemma}
\label{lem:hcscomp}
  $\forall s \in \PHl, \forall h = \PHhit{a_i}{b_j}{b_k} \in \PHh, a \in \cs, b \in \components,$\\
  $(\PHget{s}{a} = a_i \wedge \PHget{s}{b} = b_j \wedge \pfp_s(a) = a_i) \Rightarrow (\exists \delta \in \restriction{\Sce}{1}(s),
  (s \PHplay \delta) \PHPtrans (s \PHplay \delta \PHplay h))$
\end{lemma}
\begin{proof}
  Similar to the proof of \pref{lem:hcompcomp};
  as $a_i \in \pfp_s(a)$, $a_i \in (s \PHplay \delta)$.
\end{proof}

\modMF{%
\begin{example}
  Considering the model of \pref{fig:ph-livelock} and starting from the state
  $s = \PHstate{a_1, b_0, ab_{00}, z_0}$,
  \pref{lem:hcompcomp} states that the action $h = \PHfrappe{a_1}{a_1}{a_0}$
  is playable after a scenario $\delta$ of prioritized actions.
  In this case, $h$ is indeed playable in the state
  $s \play \delta = \PHstate{a_1, b_0, ab_{10}, z_0}$,
  where $\delta = (\PHfrappe{a_1}{ab_{00}}{ab_{10}}) \in \restriction{\Sce}{1}(s)$.
\end{example}
}%

\medskip

\modMF{%
This section aimed at defining the PH framework in a general fashion,
and focused on a more specific class of PH models
that will be used later in this paper.
We gave a definition of the notion of cooperative sort,
a special sort with no explicit biological existence or meaning,
which is used to model the joint action of several components
on another one.
Using this definition, we defined a specific class of PH models
that assign a specific role to the set of actions
of highest priority:
these actions can only be used to update cooperative sorts to ensure that
they are equivalent to Boolean gates.
The static analysis developed in the next section can only be applied
to PH models belonging to this specific class.
However, it has to be noted that this class is in fact as expressive as
the class of all possible PH models with priorities (see \pref{ssec:flattening}).
}%
